Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2017-12-06 article3.Rmd
Some smaller revisions

A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality. In the present study, we challenge common notions of this by taking a long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences. The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is analysed according to gender and age group (working age vs retired), testing whether the same patterns prevail in old age---the retired population---as in working age.  Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.

@@

One of the central aspects of survival is social class and access to economic
and other resources. In present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and its impact even
seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserve their appellation of
'fundamental causes' of disease and death". The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph95; @liph98]. The reigning view is that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies. This is a
reasonable assumption since these societies were often characterized by
enormous socioeconomic differences. Knowing that access to different kinds of resources
provides advantages in all aspects of life, the health advantage of higher
classes ought to be obvious.

However, recent studies investigating social inequality in health and mortality
with micro-data have questioned the generality of the assumed
pattern [@bevp11]. Solid empirical evidence about the process is lacking however and studies
focusing on the issue are still few. There is a need for additional reliable
studies from different geographical and historical settings in order to 
better understand the role of socio-economic conditions in health and survival over time.

Social conditions affect health and mortality differently during the life course. There are different hypotheses about the impact in adulthood and old age respectively. The differences either converge in old age (status levelling), are constant (status maintenance), or diverge (cumulative advantage) [@hof08]. Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors. The status maintenance hypothesis basically assumes 
continuity in the determinants of social health inequalities from adulthood to old age.
The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages tend to persist and accumulate during life in a negative spiral rewarding some while disfavouring others. This leads to larger differences in old age.

@@

The recent extension in time of data at the Demographic Data Base (DDB),
Umeå University (http://www.cedar.umu.se) by including the period 1900 to
1950 makes it possible to investigate the long-term development of social
inequality in mortality in this part of Sweden. The data for the present study come from two large
population databases at the DDB, which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden. Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time. The period 1801--1950 is covered by the database Poplink [@wee16]. Poplink is based on linked parish records, allowing us to reconstruct life biographies on people as long as they remained in the region. The records are linked within, but not
between, the regions. Data from Poplink are accessed for the period 1801--1950.

@@

Female labour is much better covered in the Linnaeus database. A
complication, however, is that it is  difficult to define households in the
same way as in Poplink. Extramarital cohabitation became common and female
labour force participation developed as the norm, in Sweden during this
period. The husband's occupation, as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable. Noti, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].

>>>>

A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality. In the present study, we challenge common notions of this by taking a long-term perspective on the development of social class inequalities in mortality in the adult and elderly population with a special focus on the relative differences. The arguments for our statements are based on an investigation of the Skellefteå and Umeå regions in northern Sweden for the periods 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is analysed according to gender and age group (working age vs retired), testing whether the same patterns prevail in old age---the retired population---as in working age.  Furthermore we place this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.

@@

One of the central aspects of survival is social class and access to economic
and other resources. In present-day welfare societies, social position, either according to social class, education or income, is a
strong determinant when it comes to health and mortality and its impact even
seems to be increasing [@kuetal04; @macketal16; @frlu07; @broba12; @strand10]. @liph02 suggest that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserve their appellation of
'fundamental causes' of disease and death". The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in their theory [@liph95; @liph98]. The reigning view is that socio-economic health and mortality inequalities were large in historical societies, probably larger than in modern societies. This is a
reasonable assumption since these societies were often characterized by
enormous class differences. Knowing that access to different kinds of resources, such as economic, social and cultural capital, provides advantages in all aspects of life, the health advantage of higher classes ought to be obvious.

However, recent studies investigating social inequality in health and mortality
with micro-data have questioned the generality of the assumed
pattern [@bevp11]. Solid empirical evidence about the process is lacking however and studies
focusing on the issue are still few. There is a need for additional reliable
studies from different geographical and historical settings in order to 
better understand the role of class and socio-economic conditions in health and survival over time.

Social position affects health and mortality differently during the life course. There are different hypotheses about the impact in adulthood and old age respectively. The differences either converge in old age (status levelling), are constant (status maintenance), or diverge (cumulative advantage) [@hof08]. Diminishing differences may be a consequence of the circumstance that biological factors become increasingly important during the ageing process and in old age, leaving less impact for social factors. The status maintenance hypothesis basically assumes 
continuity in the determinants of social health inequalities from adulthood to old age.
The cumulative (dis)advantage hypothesis [@dan03], implies that advantages and disadvantages tend to persist and accumulate during life in a negative spiral rewarding some while disfavouring others. This leads to larger differences in old age.

@@

The recent extension in time of data at the Demographic Data Base (DDB),
Umeå University (http://www.cedar.umu.se) by including the period 1900 to
1950 makes it possible to investigate the long-term development of social
inequality in mortality in this part of Sweden. The data for the present study come from two large
population databases at the DDB, which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden. Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time. The period 1801--1950 is covered by the database Poplink [@wee16]. Poplink is based on linked parish records, allowing us to reconstruct life biographies on people as long as they remained in the region. The records are linked within, but not
between, the regions.

@@

Female labour is much better covered in the Linnaeus database. A
complication, however, is that it is  difficult to define households in the
same way as in Poplink. Extramarital cohabitation became common and female
labour force participation developed as the norm, in Sweden during this
period. The husband's occupation, as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable. Notice, however, that there was strong social homogamy within married couples at this time, indicating that the results would not differ in any decisive way with the use of occupations on individual level instead of household level for defining social class based on the Linneaus data [@bjork92].
